---
title: "_Matter 2_: User guide for flexible out-of-memory data structures"
author: "Kylie Ariel Bemis"
date: "Revised: October 23, 2022"
output:
  BiocStyle::html_document:
  toc: true
vignette: >
  %\VignetteIndexEntry{1. Matter 2: User guide for flexible out-of-memory data structures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r style, echo=FALSE, results='asis'}
BiocStyle::markdown()
```

```{r setup, echo=FALSE, message=FALSE}
library(matter)
register(SerialParam())
```

# Introduction 


*Matter 2* provides an updated and more robust C++ backend to `matter` objects, along with a completely new implementation of sparse arrays and new signal processing functions for nonuniform sparse signal data.

Originally designed as a backend for the *Cardinal* package, The first version of *Matter* was constantly evolving to handle the ever-increasing demands of larger-than-memory mass spectrometry (MS) imaging experiments. While it was designed to be flexible from a user's point-of-view to handle a wide array for file structures beyond the niche of MS imaging, its codebase was becoming increasingly difficult to maintain and update.

*Matter 2* was re-written from the ground up to simplify some features that were rarely needed in practice and to provide a more robust and future-proof codebase for further improvement. Some advancements include:

- XXX

- YYY

- ZZZ



# Installation

*Matter* can be installed via the *BiocManager* package.

```{r install, eval=FALSE}
install.packages("BiocManager")
BiocManager::install("matter")
```

The same function can be used to update *Matter* and other Bioconductor packages.

Once installed, *Matter* can be loaded with `library()`:

```{r library, eval=FALSE}
library(matter)
```


# Usage

Coming soon.....

Coming soon.....

Coming soon.....



# Session information

```{r session-info}
sessionInfo()
```



